\documentclass[a4paper,10pt]{report}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{verbatim}
\usepackage{glossaries}
% \usepackage{glossary}
\usepackage{fancybox}
\usepackage[bookmarks=true]{hyperref}
% \usepackage[ps2pdf, bookmarks=true]{hyperref}

% \oddsidemargin -0.04cm   % read Lamport p.163
% \evensidemargin -0.04cm
% \textwidth 16.59cm
% \textheight 21.94cm

% \renewcommand{\baselinestretch}{2.0}
\renewcommand{\baselinestretch}{1.5}
\topmargin -0.5cm        % read Lamport p.163
\oddsidemargin 0.5cm   % read Lamport p.163
\textwidth 15cm
\textheight 23cm


% Title Page
\title{OpenVGA: An Open-Source PCI Graphics Adapter}

\author{Patrick Suggate}

% Bring in the listings style of Verilog (in the file Verilog_Def.tex)
\input{Verilog_Def}
\lstset{language=Verilog}
\lstset{breaklines=true}
\lstset{basicstyle=\ttfamily}

\makeglossary

\begin{document}
\maketitle

	
\abstract{This thesis describes the development of OpenVGA, an open-source
graphics adapter. OpenVGA uses programmable logic for its core functionality, and
features a processor logic core for data processing tasks. A novel
transport-triggered processor architecture, and a more traditional RISC
processor, were designed and evaluated for the data processing tasks within
OpenVGA.
 
OpenVGA is a free and open-source hardware project. The PCB artwork and the
source code, written in Verilog, Assembly, Python, and C/C++, can be freely
distributed and modified in accordance to the conditions of the GPL. OpenVGA was
designed to use a two-layer PCB and readily-available, low-cost ICs. The hardware
is therefore easy to fabricate by individuals wishing to participate in future
development.
 
The project produced many Wishbone-compliant logic cores including: a small,
fast, 16-bit TTA processor which operates up to 190~MHz when synthesised for a
Spartan-3 FPGA; a 16-bit RISC processor that operates at up to 140~MHz; a data
cache which has a dual-clock, 2-way set-associative architecture and operates at
\texttt{>}150 MHz; a small, fast PCI-to-Wishbone bridge that supports multiple
clock domains, by using asynchronous FIFOs; a VGA and DVI compatible
display-redraw logic core; and a SDRAM controller that operates at up to 120 MHz.
The Wishbone-interconnect standard allows all the logic cores to be used within
other open-source projects too.
 
The TTA16 processor is significantly faster than existing FPGA processor cores.
The RISC16 processor also has very good performance when compared with existing
FPGA processors. These results show that open-source, FPGA-based, graphics
adapters are feasible, and in particular that TTA processors can be used as
high-performance, small-footprint general-purpose processors within FPGAs.}


\chapter*{Acknowledgements\markboth{Acknowledgements}{Acknowledgements}}

This thesis would have been impossible without the help and support of many
people. I would like to thank my supervisor, Dr. Tim Molteno, for all his time
and enthusiasm. He contributed considerable effort and advice to keep the OpenVGA
project on-track.

I would also like to thank Dr. Neil Thomson for encouraging me to return to
university and undertake postgraduate studies. Also a thank you to all of the
other University of Otago Department of Physics staff and postgraduate students.

This project used many components developed by the open-source community and
without these OpenVGA would also have been impossible. Two pieces of software
were especially important for OpenVGA development, Icarus Verilog, developed by
Stephen Williams, and the GtkWave waveform viewer, with the project's development
led by Tony Bybell.



\tableofcontents
\listoftables
\listoffigures

% Glossary. All definitions not located elsewhere are within `defs.tex'.
\printglossary
\include{defs}


\newcommand\mmodule[6] %
{
\begin{center}
\shadowbox{
\begin{tabular}{l r r r}%
\textbf{Module:}	& \multicolumn{3}{l}{\begin{minipage}[t]{0.7\linewidth}#2\end{minipage}}\\%
\textbf{Description:}	& \multicolumn{3}{l}{\begin{minipage}[t]{0.7\linewidth}\raggedright#3\end{minipage}}\\%
\textbf{Related Files:}	& \multicolumn{3}{l}{\begin{minipage}[t]{0.7\linewidth}\raggedright#4\end{minipage}}\\%
\textbf{Testing Files:}	& \multicolumn{3}{l}{\begin{minipage}[t]{0.7\linewidth}\raggedright#5\end{minipage}}\\%
\textbf{Author:}	& #1 & \multicolumn{2}{r}{\textbf{License:} #6}\\%
\end{tabular}
}
\end{center}
}

\newcommand\ldescript[3]{
\textbf{#1:} & \multicolumn{3}{l}{
		\begin{minipage}[t]{#2\linewidth}\raggedright#3\end{minipage}
    }
}

\newcommand\filedescript[6] %
{
\begin{center} \shadowbox{
	\begin{tabular}{l r r r}%
		\ldescript{File}{0.7}{#2}	\\
		\ldescript{Description}{0.7}{#3}	\\
		\ldescript{Related Files}{0.7}{#4}	\\
		\ldescript{Testing Files}{0.7}{#5}	\\
		\textbf{Author:}	& #1 & \multicolumn{2}{r}{\textbf{License:} #6}\\
	\end{tabular}
}	\end{center}
}

\newcommand\bigdescript[2] %
{
\begin{minipage}{#1\linewidth}#2\end{minipage}
}


\newcommand\regdescript[4] %
{
	\shadowbox{
		\begin{tabular}{l r r r}%
%         	\multicolumn{4}{l}{\textbf{TTA16 Registers and Aliases}}	\\
			Functional Unit Name:	& \multicolumn{3}{l}{
				\begin{minipage}[t]{0.4\linewidth}\texttt{#1}\end{minipage}}\\%
			Trigger register(s) and aliases:	& \multicolumn{3}{l}{
				\begin{minipage}[t]{0.4\linewidth}\texttt{#2}\end{minipage}}\\%
			Operand register(s) and aliases:	& \multicolumn{3}{l}{
				\begin{minipage}[t]{0.3\linewidth}\texttt{#3}\end{minipage}}\\%
			Result Register(s):	& \multicolumn{3}{l}{
				\begin{minipage}[t]{0.3\linewidth}\texttt{#4}\end{minipage}}\\%
		\end{tabular}
	}
}


\include{intro}
\include{background}
\include{openvga}

% Three chapters on the internal architecture.
\include{cpu}
\include{memory}
\include{io}

% Errata, progress
\include{conclusions}


\bibliographystyle{plain}
\bibliography{thesis_bib}

\appendix
\include{appendices}
\include{wishbone}
\include{tta_prog}
\include{risc_prog}


\end{document}
